The Linear Instability of Dilute Ultrarelativistic \epm\ Pair Beams

### 2016-07-26 oblique.tex
initial fixed of Avery's comments

\newcommand{\betaMEAN}{\ensuremath{\beta_{\rm M}}}
\newcommand{\epm}{\ensuremath{e^+e^-}}

@@

\begin{abstract}
The annihilation of TeV photons from extragalactic TeV sources and the extragalactic background light produces ultrarelativistic $\epm$ beams, which are subject to powerful plasma instabilities that sap their kinetic energy.
Here we study the initial conditions of these \epm\ \ab{[Should this be $e^+$-$e^-$ or $e^\pm$?]} beams and the linear phase of the plasma instabilities that cool them.  Starting from an isotropic distribution in the center of mass frame, we show that the velocity dispersions parallel and transverse to the direction of beam propagation become highly anisotropic under Lorentz boosts and scale differently with the Lorentz factor, $\gamma$.  Using these results to construct a simplified distribution function, we then calculate the linear growth rate of the beam plasma and oblique instability in both the reactive and kinetic regimes.  We reproduce the well-known reactive growth rates for both the beam plasma and oblique mode and show that they both have similar growth rates in the kinetic regime.  In contrast to the earlier claim by the authors that the oblique mode is dominant, we instead find that both the beam-plasma (parallel) and oblique modes are similarly strong.  In addition, we also delineate the conditions for applicability for the reactive and kinetic regimes and find that the beam plasma mode transitions to the reactive regime at a lower $\gamma$ than the oblique mode due to a combination of their different scalings and the anisotropy of the velocity dispersions.  Applying these results to the ultrarelativistic $\epm$ beams from TeV blazars, we \ab{\sout{not only}?} confirm that these beams are unstable to the kinetic oblique mode.  For sufficiently cold beams in the beam direction or for sufficiently larger Lorentz factor, these beams are also unstable to the reactive beam-plasma instability.  We show that overall growth rates for both the oblique and beam plasma growth rates are similar and in line with our earlier claims.  These results strengthen the case that powerful plasma instabilities  sap the energy of the ultrarelativistic $\epm$ beams as they propagate through intergalactic space, which has profound consequences on constraints of the intergalactic magnetic field, the thermal state of the intergalactic medium, and the formation of structure in the universe.
\ab{[The only issue with the above is that the primary issues people have raised are with nonlinear effects, e.g., saturation or nonlinear coupling to existing background structures.  Should we be quite as strident then in the concluding sentences?]}
\end{abstract}

@@

The \Fermi satellite and ground-based imaging atmospheric Cerenkov \ab{[Cherenkov?  I think MAGIC, VERITAS, and HESS all use the ``h'' on their websites.]} telescopes such as
% Phil: I added "imaging atmospheric" to distinguish this class of
% telescopes from imaging water CTs, e.g., MILAGRO; I am not aware of
% any space based IACT apart from a planned telescope on the ISS, but
% who knows whether this will ever fly
H.E.S.S., MAGIC, and VERITAS\footnote{High Energy
  Stereoscopic System, Major Atmospheric Gamma Imaging Cerenkov Telescope, Very
  Energetic Radiation Imaging Telescope Array System.} have demonstrated that the high
energy Universe is teeming with energetic very high-energy gamma-ray 
(VHEGR, $E > 100\,\GeV$) sources, the extragalactic component of which mainly consists of TeV
blazars with a minority population of other sources  such as radio
and starburst galaxies. These extragalactic VHEGR emitters produce TeV photons that are greatly attenuated via annihilation upon soft photons in the extragalactic background light
(EBL) and produce pairs \citep[see, e.g.,][]{Goul-Schr:67,Sala-Stec:98,Nero-Semi:09}.

@@

These two important conclusions depend on IC cooling dominating the evolution of the ultra-relativistic pairs.  However, it was recently found that plasma instabilities driven by the
ultrarelativistic pair beams  likely are the dominant cooling mechanisms
\citep{paperI,Schlickeiser+12,Schlickeiser+13}, depositing this energy
as heat in the intergalactic medium \citep{paperII,paperIII}.  Therefore, the lack of an observed
 IC halo emission from TeV blazars does not imply the existence of the IGMF as previous groups have argued
\citep{paperI,Schlickeiser+12,Schlickeiser+13}.

@@

To this end, a full nonlinear study is called for.  We begin this task in this initial work where we study the distribution function of the $e^+e^-$ pairs that are produced from VHEGR-EBL photon annihilation.  We study the evolution of a distribution function under Lorentz transformations to develop an analytic understanding of how the perpendicular and parallel velocity dispersions transform under boosts.  Using this understanding, we then develop a simple description of the distribution function of the beam, which we then use to calculate the unstable modes analytically in both the reactive (hydrodynamic) and kinetic regimes.

\ab{[New paragraph]}
Here the reactive instability refers to the instability where the entire beam participates in the instability.  In particular, all the beam particles are resonant with the unstable wave on a timescale longer than the growth time of the instability.  The reactive instability is also referred to as the hydrodynamic instability since the instability can be derived from the fluid equations instead of kinetic theory.  On the other hand, in the kinetic regime, only a fraction of the beam particles are resonant with the beam over a the growth time of the instability, which reduces the growth rate compared to the reactive instability for the same beam density and beam Lorentz factor.  We recover the well-known results for the reactive regime for both the beam-plasma and oblique modes.  We also derive the growth rate for these two instabilities in the kinetic regime and delineate the range of applicability for both the reactive and kinetic cases and apply these results for ultrarelativistic \epm\ pair beams.

This paper is organized as follows.  In Section \ref{sec:setup}, we describe the transformation properties of an ultrarelativistic \epm\ beam in terms of its distribution function.  We then calculate the various linear instabilities that this beam is subject to in section \ref{sec:linear}.  In particular, we pay careful attention to both the reactive (or hydrodynamic) and kinetic regimes of the beam plasma and oblique instabilities and the transition between the two.  Applying these results to TeV $e^+e^-$ pair beams that arise from TeV photon pair production in Section \ref{sec:application},
we demonstrate that despite the extraordinary coldness of the beam we are always in the kinetic regime for the oblique mode, but may be in the reactive regime for the beam plasma mode.  However, for the relevant parameters, the growth rates calculated in either regime are similar.  We close with a discussion the implications of this work and application of these results for nonlinear theory in Section \ref{sec:conclusions}.

@@

As stated in the Introduction, VHEGR photons pair produce on encountering EBL
photons as they propagate throughout the universe \citep{Gould+66}, and this
attenuation of VHEGR flux has been used as a probe of the EBL
\citep{Stec-deJa-Sala:92,deJa-Stec-Sala:94,Ahar_etal:06}.  The basic requirement
of this process is that the energies of the VHEGR ($E_{\rm ph}$) and the EBL
photon ($E_{\rm ebl}$) exceed the rest mass energy of the $e^+e^-$ pair in the
center of momentum (COM) frame, i.e., $2 E E_{\rm ebl}(1-\cos\theta) \geq 4
m_e^2 c^4$, where $\theta$ is the relative angle of propagation in the lab
frame. As a result, an $e^\pm$ \ab{[\epm elsewhere]} pair can be produced with Lorentz factor $\gamma = \left(1-v^2/c^2\right)^{-1/2} \simeq E/2m_e
c^2$, where $v$ is the velocity of the pairs \citep{Goul-Schr:67}.  Here, we discuss the distribution function of the
pair beam that emerges from this process.

@@

We now adopt perturbations of the form $\delta \sim \exp\left(i \kvec\cdot\rvec - i \omega t\right)$ and without loss of generality assume that $\kvec = (k_x, 0, k_z)$, where $k_z$ is along the beam direction.  Linearizing the Vlasov-Maxwell equations then leads to the dispersion relation:
%\begin{equation}\label{eq:dispersion relation}
% \epsilon = 1 - \frac{\omega_{p}^2}{k}\int \frac{\kvec \cdot \gradp F}{\omega - \kvec\cdot\vel} d^3\!p = 0,
%\end{equation}
\begin{equation}\label{eq:dispersion relation}
 \epsilon = 1 + \sum_s \frac{m_e \omega_{p,s}^2}{k^2}\int \frac{\kvec \cdot \gradp F_s}{\omega - \kvec\cdot\vel} d^3\!p = 0,
\end{equation}
where $\epsilon$ is the simplified dielectric function, and for each species $\omega_{p,s}^2\equiv 4\pi e^2 n_s/m_e$ is the plasma frequency, $n_s\equiv\int f_{0s} d^3\!p$ is the number density, and $F_s\equiv f_{0s} / n_s$ is the normalized \ab{background} distribution function.
Upon integrating by parts, Equation (\ref{eq:dispersion relation}) becomes
%\begin{eqnarray}
% 1 - \frac{4\pi q_s^2}{m_s k^2}\int F_s\kvec \cdot \gradp\frac{1}{\omega - \kvec\cdot\vel} d^3p &=& 0\\
% 1 - \frac{4\pi q_s^2}{m_s k^2}\int F_s \frac{k^2 - (\kvec\cdot\betavec)^2}{\gamma(\omega - \kvec\cdot\vel)^2} d^3p &=& 0. \label{eq:dispersion relation by parts}
%\end{eqnarray}
\begin{multline}
  \epsilon = 
  1 - \sum_s \frac{m_e\omega_{p,s}^2}{k^2} \int F_s\kvec \cdot \gradp\frac{1}{\omega - \kvec\cdot\vel} d^3p\\ 
  =
  1 - \sum_s \frac{\omega_{p,s}^2}{k^2c^2}\int F_s \frac{k^2c^2 - (\kvec\cdot\vel)^2}{\gamma(\omega - \kvec\cdot\vel)^2} d^3p 
  =
  0\,.
  \label{eq:dispersion relation by parts}
\end{multline}
%\ac{[$\beta$ vs. $v$ issue above.]}

There are two distinct, often qualitatively different, regimes in which we may
consider the implications of this dispersion relation.  The first is the cold
plasma limit or the hydrodynamic or reactive limit.  The hydrodynamic limit is
aptly named because the resulting dispersion relation that is found could have
also been calculated directly from the continuity equation and the momentum
equation.  In this limit, the internal distribution of the particles of the
background or beam are irrelevant to the physics of the instability and it is
only the bulk response that is important.  In particular, this means that the
beam particles are \ab{resonant \sout{resonance}} with the unstable wave over a timescale much longer
than the growth time \ab{\sout{as stated above}}, i.e., the beam particles
do not drift a distance larger than the wavelength of the unstable mode over the
growth time of the instability.  The second is the kinetic regime, where the
internal distribution of  beam particles is
important to the physics of the instability. Here, only a fraction of
 beam particles stay within one wavelength of the unstable mode
over the growth time of the instability.  Moreover, the bulk of
the plasma (background or beam) does not respond to the disturbance\ab{; instead, \sout{. Instead,}}
only a fraction of  particles is relevant for
driving (instability) or damping (Landau damping).  We discuss below the
evaluation of the dispersion relation in these two regimes, which gives two
regimes of instability, and the delineation between \ab{them. \sout{these two regimes.}}

@@

Starting with the dispersion relation (\ref{eq:dispersion relation by parts}), we first consider the instability of a cold plasma beam. 
Taking the limit of Equation (\ref{eq:distribution function beam}) as $k_BT_t  \rightarrow 0$, for a target plasma $v_0=0$ and a beam plasma $v_0=\vbeam$\footnote{That is, we set $F_s(\bmath{p})=\delta^3(\bmath{p}-\bmath{p}_{0s})$ where $\bmath{p}_{0s}\equiv \gamma_0 m_e v_0 \hat{\bmath{z}}$ is the momentum associated with $v_0$.}, we find
\begin{equation}\label{eq:dispersion relation reactive}
 1 - \frac{\omega_{p,t}^2}{\omega^2} - \frac{\omega_{p,b}^2}{\gamma^3(\omega-k_z \vbeam)^2}\frac{\gamma^2 k_x^2 + k_z^2}{k_x^2 + k_z^2} = 0.
\end{equation}
For $k_x = 0$, we recover the same beam-plasma instability which was described in the Appendix of \citet{paperI}.

@@

For $k_x = 0 \rightarrow Z_x = 0$, this reduces to the beam-plasma growth rate, which is
\begin{equation}\label{eq:growth rate bp}
 \Gamma = \Gamma_{\rm TS} \equiv \frac{\sqrt{3}}{2^{4/3}}\left(\frac{n_b}{n_t}\right)^{1/3}\frac{\omega_{p,t}}{\gamma},
\end{equation}
which we denote the beam-plasma or ``two-stream'' growth rate.
For the more general case where $Z_x\ne 0$, this becomes the oblique instability studied by \citet{Bret-Grem-Beni:10}  Indeed for $\gamma \gg 1$ and \ab{$Z_x\gg1$ \sout{$Z_x \rightarrow \infty$}}, the growth rate approaches oblique growth rate:
\begin{equation}
 \Gamma = \Gamma_{\rm ob} \equiv \frac{\sqrt{3}}{2^{4/3}}\left(\frac{n_b}{n_t}\right)^{1/3}\frac{\omega_{p,t}}{\gamma^{1/3}},
\end{equation}
which is much faster than the beam-plasma growth rate, $\Gamma_{\rm TS}$

@@

In a continuous system, these waves do not matter in comparison to the unstable mode at $k\lambda_D = 1$.   However, for discrete numerical systems, which do not sufficiently resolve the most unstable modes, these \ab{sub-maximal \sout{sub maximal}} modes drive the growth of the instability of numerically calculated beam-plasma systems, which may lead to an incorrect nonlinear state in comparison to the physical system.

@@

The growth rate expressed in Equation (\ref{eq:growth rate reactive}) is in the reactive (or hydrodynamic) regime as the dispersion relation (Equation (\ref{eq:dispersion reactive})) could have been derived from the fluid equations.  Here all the particles participate in the instability.  However, kinetic theory \ab{marks \sout{masks}} another regime of the instability, where only a fraction of the particle participate in the instability, i.e., the kinetic regime.  We now derive the growth rate of the instability in the kinetic regime.

We begin first with the distribution function for the target plasma:
\begin{equation}
 F_t = \left(\frac{1}{2\pi m_e k_B T_t}\right)^{3/2}\exp\left(-\frac{p^2}{2m_e k_B T_t}\right),
\end{equation}
where the target plasma is assumed to be nonrelativistic, $\bmath{p}=m_e\vel$ is the nonrelativistic momentum,  and $T_t$ is the temperature of the target background plasma.
For the beam plasma, we \ab{again} adopt the Maxwell-J{\"u}ttner distribution (\ab{Equation \sout{eq.}}~(\ref{eq:distribution function beam})).
%\begin{equation}\label{eq:distribution function beam}
%F_b = \frac{n_b}{4\pi\gammabeam m^2 T K_2(m/T_{\rm L})} \exp\left(-\frac {\gammabeam(E - \betabeam \cdot \pmom)} {T}\right),
%\end{equation}
Inserting these into the dispersion relation\ab{, Equation} (\ref{eq:dispersion relation}), we find
%\begin{equation}\label{eq:dispersion kinetic}
% 1 - \frac{\omega_{p,t}^2}{k}\int F_t\frac{k^2 - (\kvec\cdot\betavec)^2}{\gamma(\omega - \kvec\cdot\vel)^2} d^3p - \frac{\omega_{p,t}^2}{k}\int \frac{\kvec\cdot\gradp F_b}{\omega - \kvec\cdot\vel}d^3p  = 0,
%\end{equation}
\begin{equation}\label{eq:dispersion kinetic}
 1 - \frac{\omega_{p,t}^2}{k^2c^2}\int F_t\frac{k^2c^2 - (\kvec\cdot\vel)^2}{\gamma(\omega - \kvec\cdot\vel)^2} d^3p + \frac{m_e\omega_{p,b}^2}{k^2}\int \frac{\kvec\cdot\gradp F_b}{\omega - \kvec\cdot\vel}d^3p  = 0,
\end{equation}
where we have integrated by parts only the second term, associated with target plasma.

@@

In Paper I \ab{[Paper I is nowhere defined]}, we compared the cold plasma growth rates of the oblique and beam-plasma instabilities and noted that the oblique cold growth rate is larger.  While we also noted \ab{that the} oblique instability was in the kinetic regime in Paper I, we made no effort to study the regime of instability of the beam-plasma case.  Here we have shown that while the oblique growth rate is kinetic, the beam-plasma rate is marginally reactive.  We have also shown that the kinetic growth rates for both the oblique and beam-plasma instabilities are similar.  As a result, both instabilities (beam-plasma and oblique) have similar growth rates for the case of  pair beams from TeV blazers. \ab{[Where did we show that the beam-plasma instability is marginally reactive? Does this belong in the conclusions?]}

@@

To begin let us return to the reactive instability.  For the growth rate of the reactive instability \ab{in Equation} (\ref{eq:growth rate reactive}) to be valid, the velocity dispersion \ab{must be vanishingly small. Quantitatively, this limit requires \sout{$\rightarrow 0$.   To reach this limit, we demand}}
\begin{equation}
 \kvec\cdot\Delta \vel \ll \omega - \kvec\cdot\vel_b.
\end{equation}
Taking $\kvec$ to be $\omega_{p,t}/\vel_b$, so that the RHS is $\approx \Gamma$, we find
\ab{[While it is true that $\omega-\omega_{p,t}\approx\Gamma$, this is not fully fleshed out in the main text; rather it is an inline comment in Appendix B.  Thus, this is likely to be a confusing part.  Also, if $\kvec=\omega_{p,t}/\vel_b$ then isn't hte LHS $\Delta v_\parallel/v_b$.]}
\begin{equation}\label{eq:oblique reactive regime}
 \frac{\Delta v_{\perp}}{\vbeam} \ll \left(\frac{n_b}{\gammabeam n_t}\right)^{1/3}
%\left(1 - \beta_b^2\cos^2\Psi\right)^{1/3}, 
\end{equation}
where we have dropped constant factors of order unity and assumed that $Z_x \sim O(1)$.  Hence, this defines the upper limit on the velocity dispersion of the plasma for the cold-plasma approximation to hold and, hence, the range of validity for the reactive oblique growth rate \ab{Equation} (\ref{eq:growth rate reactive}).  For $Z_x \ll \gamma^{-2}$, we recover the condition for the relativistic, reactive beam-plasma instability:
\begin{equation}\label{eq:bp reactive regime}
\frac{\Delta v_{\parallel}}{\vbeam} \ll \gammabeam^{-1}\left(\frac{n_b}{n_t}\right)^{1/3}.
\end{equation}

>>>>

\newcommand{\betaMEAN}{\ensuremath{\beta_{\rm M}}}
\newcommand{\epm}{\ensuremath{e^{\pm}}}

@@

\begin{abstract}
The annihilation of TeV photons from extragalactic TeV sources and the extragalactic background light produces ultrarelativistic $\epm$ beams, which are subject to powerful plasma instabilities that sap their kinetic energy.
Here we study the initial conditions of these \epm\ beams and the linear phase of the plasma instabilities that cool them.  Starting from an isotropic distribution in the center of mass frame, we show that the velocity dispersions parallel and transverse to the direction of beam propagation become highly anisotropic under Lorentz boosts and scale differently with the Lorentz factor, $\gamma$.  Using these results to construct a simplified distribution function, we then calculate the linear growth rate of the beam plasma and oblique instability in both the reactive and kinetic regimes.  We reproduce the well-known reactive growth rates for both the beam plasma and oblique mode and show that they both have similar growth rates in the kinetic regime.  In contrast to the earlier claim by the authors that the oblique mode is dominant, we instead find that both the beam-plasma (parallel) and oblique modes are similarly strong.  In addition, we also delineate the conditions for applicability for the reactive and kinetic regimes and find that the beam plasma mode transitions to the reactive regime at a lower $\gamma$ than the oblique mode due to a combination of their different scalings and the anisotropy of the velocity dispersions.  Applying these results to the ultrarelativistic $\epm$ beams from TeV blazars, we confirm that these beams are unstable to the kinetic oblique mode.  For sufficiently cold beams in the beam direction or for sufficiently larger Lorentz factor, these beams are also unstable to the reactive beam-plasma instability.  We show that overall growth rates for both the oblique and beam plasma growth rates are similar and in line with our earlier claims.  These results strengthen the case that powerful plasma instabilities  sap the energy of the ultrarelativistic $\epm$ beams as they propagate through intergalactic space, which has profound consequences on constraints of the intergalactic magnetic field, the thermal state of the intergalactic medium, and the formation of structure in the universe.
\ab{[The only issue with the above is that the primary issues people have raised are with nonlinear effects, e.g., saturation or nonlinear coupling to existing background structures.  Should we be quite as strident then in the concluding sentences?]}
\end{abstract}

@@

The \Fermi satellite and ground-based imaging atmospheric Cherenkov telescopes such as
% Phil: I added "imaging atmospheric" to distinguish this class of
% telescopes from imaging water CTs, e.g., MILAGRO; I am not aware of
% any space based IACT apart from a planned telescope on the ISS, but
% who knows whether this will ever fly
H.E.S.S., MAGIC, and VERITAS\footnote{High Energy
  Stereoscopic System, Major Atmospheric Gamma Imaging Cerenkov Telescope, Very
  Energetic Radiation Imaging Telescope Array System.} have demonstrated that the high
energy Universe is teeming with energetic very high-energy gamma-ray 
(VHEGR, $E > 100\,\GeV$) sources, the extragalactic component of which mainly consists of TeV
blazars with a minority population of other sources  such as radio
and starburst galaxies. These extragalactic VHEGR emitters produce TeV photons that are greatly attenuated via annihilation upon soft photons in the extragalactic background light
(EBL) and produce pairs \citep[see, e.g.,][]{Goul-Schr:67,Sala-Stec:98,Nero-Semi:09}.

@@

These two important conclusions depend on IC cooling dominating the evolution of the ultra-relativistic pairs.  However, it was recently found that plasma instabilities driven by the
ultrarelativistic pair beams  likely are the dominant cooling mechanisms
(\citealt{paperI}, hereafter ACP12, \citealt{Schlickeiser+12,Schlickeiser+13}), depositing this energy
as heat in the intergalactic medium \citep{paperII,paperIII}.  Therefore, the lack of an observed
 IC halo emission from TeV blazars does not imply the existence of the IGMF as previous groups have argued
(ACP12;\citealt{Schlickeiser+12,Schlickeiser+13}).

@@

To this end, a full nonlinear study is called for.  We begin this task in this initial work where we study the distribution function of the $\epm$ pairs that are produced from VHEGR-EBL photon annihilation.  We study the evolution of a distribution function under Lorentz transformations to develop an analytic understanding of how the perpendicular and parallel velocity dispersions transform under boosts.  Using this understanding, we then develop a simple description of the distribution function of the beam, which we then use to calculate the unstable modes analytically in both the reactive (hydrodynamic) and kinetic regimes.

Here the reactive instability refers to the instability where the entire beam participates in the instability.  In particular, all the beam particles are resonant with the unstable wave on a timescale longer than the growth time of the instability.  The reactive instability is also referred to as the hydrodynamic instability since the instability can be derived from the fluid equations instead of kinetic theory.  On the other hand, in the kinetic regime, only a fraction of the beam particles are resonant with the beam over a the growth time of the instability, which reduces the growth rate compared to the reactive instability for the same beam density and beam Lorentz factor.  We recover the well-known results for the reactive regime for both the beam-plasma and oblique modes.  We also derive the growth rate for these two instabilities in the kinetic regime and delineate the range of applicability for both the reactive and kinetic cases and apply these results for ultrarelativistic \epm\ pair beams.

This paper is organized as follows.  In Section \ref{sec:setup}, we describe the transformation properties of an ultrarelativistic \epm\ beam in terms of its distribution function.  We then calculate the various linear instabilities that this beam is subject to in section \ref{sec:linear}.  In particular, we pay careful attention to both the reactive (or hydrodynamic) and kinetic regimes of the beam plasma and oblique instabilities and the transition between the two.  Applying these results to TeV $\epm$ pair beams that arise from TeV photon pair production in Section \ref{sec:application},
we demonstrate that despite the extraordinary coldness of the beam we are always in the kinetic regime for the oblique mode, but may be in the reactive regime for the beam plasma mode.  However, for the relevant parameters, the growth rates calculated in either regime are similar.  We close with a discussion the implications of this work and application of these results for nonlinear theory in Section \ref{sec:conclusions}.

@@

As stated in the Introduction, VHEGR photons pair produce on encountering EBL
photons as they propagate throughout the universe \citep{Gould+66}, and this
attenuation of VHEGR flux has been used as a probe of the EBL
\citep{Stec-deJa-Sala:92,deJa-Stec-Sala:94,Ahar_etal:06}.  The basic requirement
of this process is that the energies of the VHEGR ($E_{\rm ph}$) and the EBL
photon ($E_{\rm ebl}$) exceed the rest mass energy of the $\epm$ pair in the
center of momentum (COM) frame, i.e., $2 E E_{\rm ebl}(1-\cos\theta) \geq 4
m_e^2 c^4$, where $\theta$ is the relative angle of propagation in the lab
frame. As a result, an $\epm$ pair can be produced with Lorentz factor $\gamma = \left(1-v^2/c^2\right)^{-1/2} \simeq E/2m_e
c^2$, where $v$ is the velocity of the pairs \citep{Goul-Schr:67}.  Here, we discuss the distribution function of the
pair beam that emerges from this process.

@@

We now adopt perturbations of the form $\delta \sim \exp\left(i \kvec\cdot\rvec - i \omega t\right)$ and without loss of generality assume that $\kvec = (k_x, 0, k_z)$, where $k_z$ is along the beam direction.  Linearizing the Vlasov-Maxwell equations then leads to the dispersion relation:
%\begin{equation}\label{eq:dispersion relation}
% \epsilon = 1 - \frac{\omega_{p}^2}{k}\int \frac{\kvec \cdot \gradp F}{\omega - \kvec\cdot\vel} d^3\!p = 0,
%\end{equation}
\begin{equation}\label{eq:dispersion relation}
 \epsilon = 1 + \sum_s \frac{m_e \omega_{p,s}^2}{k^2}\int \frac{\kvec \cdot \gradp F_s}{\omega - \kvec\cdot\vel} d^3\!p = 0,
\end{equation}
where $\epsilon$ is the simplified dielectric function, and for each species $\omega_{p,s}^2\equiv 4\pi e^2 n_s/m_e$ is the plasma frequency, $n_s\equiv\int f_{0s} d^3\!p$ is the number density, and $F_s\equiv f_{0s} / n_s$ is the normalized background distribution function.
Upon integrating by parts, Equation (\ref{eq:dispersion relation}) becomes
%\begin{eqnarray}
% 1 - \frac{4\pi q_s^2}{m_s k^2}\int F_s\kvec \cdot \gradp\frac{1}{\omega - \kvec\cdot\vel} d^3p &=& 0\\
% 1 - \frac{4\pi q_s^2}{m_s k^2}\int F_s \frac{k^2 - (\kvec\cdot\betavec)^2}{\gamma(\omega - \kvec\cdot\vel)^2} d^3p &=& 0. \label{eq:dispersion relation by parts}
%\end{eqnarray}
\begin{multline}
  \epsilon = 
  1 - \sum_s \frac{m_e\omega_{p,s}^2}{k^2} \int F_s\kvec \cdot \gradp\frac{1}{\omega - \kvec\cdot\vel} d^3p\\ 
  =
  1 - \sum_s \frac{\omega_{p,s}^2}{k^2c^2}\int F_s \frac{k^2c^2 - (\kvec\cdot\vel)^2}{\gamma(\omega - \kvec\cdot\vel)^2} d^3p 
  =
  0\,.
  \label{eq:dispersion relation by parts}
\end{multline}
%\ac{[$\beta$ vs. $v$ issue above.]}

There are two distinct, often qualitatively different, regimes in which we may
consider the implications of this dispersion relation.  The first is the cold
plasma limit or the hydrodynamic or reactive limit.  The hydrodynamic limit is
aptly named because the resulting dispersion relation that is found could have
also been calculated directly from the continuity equation and the momentum
equation.  In this limit, the internal distribution of the particles of the
background or beam are irrelevant to the physics of the instability and it is
only the bulk response that is important.  In particular, this means that the
beam particles are resonant with the unstable wave over a timescale much longer
than the growth time, i.e., the beam particles
do not drift a distance larger than the wavelength of the unstable mode over the
growth time of the instability.  The second is the kinetic regime, where the
internal distribution of  beam particles is
important to the physics of the instability. Here, only a fraction of
 beam particles stay within one wavelength of the unstable mode
over the growth time of the instability.  Moreover, the bulk of
the plasma (background or beam) does not respond to the disturbance; instead,
only a fraction of  particles is relevant for
driving (instability) or damping (Landau damping).  We discuss below the
evaluation of the dispersion relation in these two regimes, which gives two
regimes of instability, and the delineation between them.

@@

Starting with the dispersion relation (\ref{eq:dispersion relation by parts}), we first consider the instability of a cold plasma beam. 
Taking the limit of Equation (\ref{eq:distribution function beam}) as $k_BT_t  \rightarrow 0$, for a target plasma $v_0=0$ and a beam plasma $v_0=\vbeam$\footnote{That is, we set $F_s(\bmath{p})=\delta^3(\bmath{p}-\bmath{p}_{0s})$ where $\bmath{p}_{0s}\equiv \gamma_0 m_e v_0 \hat{\bmath{z}}$ is the momentum associated with $v_0$.}, we find
\begin{equation}\label{eq:dispersion relation reactive}
 1 - \frac{\omega_{p,t}^2}{\omega^2} - \frac{\omega_{p,b}^2}{\gamma^3(\omega-k_z \vbeam)^2}\frac{\gamma^2 k_x^2 + k_z^2}{k_x^2 + k_z^2} = 0.
\end{equation}
For $k_x = 0$, we recover the same beam-plasma instability which was described in the Appendix of ACP12.

@@

For $k_x = 0 \rightarrow Z_x = 0$, this reduces to the beam-plasma growth rate, which is
\begin{equation}\label{eq:growth rate bp}
 \Gamma = \Gamma_{\rm TS} \equiv \frac{\sqrt{3}}{2^{4/3}}\left(\frac{n_b}{n_t}\right)^{1/3}\frac{\omega_{p,t}}{\gamma},
\end{equation}
which we denote the beam-plasma or ``two-stream'' growth rate.
For the more general case where $Z_x\ne 0$, this becomes the oblique instability studied by \citet{Bret-Grem-Beni:10}  Indeed for $\gamma \gg 1$ and $Z_x\gg1$, the growth rate approaches oblique growth rate:
\begin{equation}
 \Gamma = \Gamma_{\rm ob} \equiv \frac{\sqrt{3}}{2^{4/3}}\left(\frac{n_b}{n_t}\right)^{1/3}\frac{\omega_{p,t}}{\gamma^{1/3}},
\end{equation}
which is much faster than the beam-plasma growth rate, $\Gamma_{\rm TS}$

@@

In a continuous system, these waves do not matter in comparison to the unstable mode at $k\lambda_D = 1$.   However, for discrete numerical systems, which do not sufficiently resolve the most unstable modes, these sub-maximal modes drive the growth of the instability of numerically calculated beam-plasma systems, which may lead to an incorrect nonlinear state in comparison to the physical system.

@@

The growth rate expressed in Equation (\ref{eq:growth rate reactive}) is in the reactive (or hydrodynamic) regime as the dispersion relation (Equation (\ref{eq:dispersion reactive})) could have been derived from the fluid equations.  Here all the particles participate in the instability.  However, kinetic theory marks another regime of the instability, where only a fraction of the particle participate in the instability, i.e., the kinetic regime.  We now derive the growth rate of the instability in the kinetic regime.

We begin first with the distribution function for the target plasma:
\begin{equation}
 F_t = \left(\frac{1}{2\pi m_e k_B T_t}\right)^{3/2}\exp\left(-\frac{p^2}{2m_e k_B T_t}\right),
\end{equation}
where the target plasma is assumed to be nonrelativistic, $\bmath{p}=m_e\vel$ is the nonrelativistic momentum,  and $T_t$ is the temperature of the target background plasma.
For the beam plasma, we again adopt the Maxwell-J{\"u}ttner distribution (Equation~(\ref{eq:distribution function beam})).
%\begin{equation}\label{eq:distribution function beam}
%F_b = \frac{n_b}{4\pi\gammabeam m^2 T K_2(m/T_{\rm L})} \exp\left(-\frac {\gammabeam(E - \betabeam \cdot \pmom)} {T}\right),
%\end{equation}
Inserting these into the dispersion relation (Equation \ref{eq:dispersion relation}), we find
%\begin{equation}\label{eq:dispersion kinetic}
% 1 - \frac{\omega_{p,t}^2}{k}\int F_t\frac{k^2 - (\kvec\cdot\betavec)^2}{\gamma(\omega - \kvec\cdot\vel)^2} d^3p - \frac{\omega_{p,t}^2}{k}\int \frac{\kvec\cdot\gradp F_b}{\omega - \kvec\cdot\vel}d^3p  = 0,
%\end{equation}
\begin{equation}\label{eq:dispersion kinetic}
 1 - \frac{\omega_{p,t}^2}{k^2c^2}\int F_t\frac{k^2c^2 - (\kvec\cdot\vel)^2}{\gamma(\omega - \kvec\cdot\vel)^2} d^3p + \frac{m_e\omega_{p,b}^2}{k^2}\int \frac{\kvec\cdot\gradp F_b}{\omega - \kvec\cdot\vel}d^3p  = 0,
\end{equation}
where we have integrated by parts only the second term, associated with target plasma.

@@

In ACP12, we compared the cold plasma growth rates of the oblique and beam-plasma instabilities and noted that the oblique cold growth rate is larger.  While we also noted \ab{that the} oblique instability was in the kinetic regime in ACP12, we made no effort to study the regime of instability of the beam-plasma case.  Here we have shown that while the oblique growth rate is kinetic, the beam-plasma rate is marginally reactive.  We have also shown that the kinetic growth rates for both the oblique and beam-plasma instabilities are similar.  As a result, both instabilities (beam-plasma and oblique) have similar growth rates for the case of  pair beams from TeV blazers. \ab{[Where did we show that the beam-plasma instability is marginally reactive? Does this belong in the conclusions?]}

@@

To begin let us return to the reactive instability.  For the growth rate of the reactive instability in Equation (\ref{eq:growth rate reactive}) to be valid, the velocity dispersion must be vanishingly small. Quantitatively, this limit requires
\begin{equation}
 \kvec\cdot\Delta \vel \ll \omega - \kvec\cdot\vel_b.
\end{equation}
Taking $\kvec$ to be $\omega_{p,t}/\vel_b$, so that the RHS is $\approx \Gamma$, we find
\ab{[While it is true that $\omega-\omega_{p,t}\approx\Gamma$, this is not fully fleshed out in the main text; rather it is an inline comment in Appendix B.  Thus, this is likely to be a confusing part.  Also, if $\kvec=\omega_{p,t}/\vel_b$ then isn't hte LHS $\Delta v_\parallel/v_b$.]}
\begin{equation}\label{eq:oblique reactive regime}
 \frac{\Delta v_{\perp}}{\vbeam} \ll \left(\frac{n_b}{\gammabeam n_t}\right)^{1/3}
%\left(1 - \beta_b^2\cos^2\Psi\right)^{1/3}, 
\end{equation}
where we have dropped constant factors of order unity and assumed that $Z_x \sim O(1)$.  Hence, this defines the upper limit on the velocity dispersion of the plasma for the cold-plasma approximation to hold and, hence, the range of validity for the reactive oblique growth rate Equation (\ref{eq:growth rate reactive}).  For $Z_x \ll \gamma^{-2}$, we recover the condition for the relativistic, reactive beam-plasma instability:
\begin{equation}\label{eq:bp reactive regime}
\frac{\Delta v_{\parallel}}{\vbeam} \ll \gammabeam^{-1}\left(\frac{n_b}{n_t}\right)^{1/3}.
\end{equation}
